\documentclass[letterpaper]{article}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}

\title{Support Density Matters: Density-Guided Feature Tracing for Training-Free Open-Vocabulary Semantic Segmentation}
\author{Anonymous Authors}

\begin{document}
\maketitle

\begin{abstract}
Training-free open-vocabulary semantic segmentation benefits from frozen
vision-language models, but dense prediction remains unstable when patch-level
visual evidence is unevenly distributed. We observe that many errors arise not
only from semantic mismatch, but from support density imbalance: boundary
patches, thin structures, and small objects often lack enough reliable
neighbors on the visual token manifold. To address this issue, we propose
\textbf{DeFT}, Density-guided Feature Tracing, a deterministic and
training-free framework. DeFT estimates token support density, traces reliable
graph and local evidence to low-support regions, and applies
structure-conserving multi-scale recalibration before text-image similarity.
Experiments follow the same training-free open-vocabulary segmentation
protocol and use a frozen CLIP-based runtime. The final DeFT row in the tables
should be filled after running the provided scripts.
\end{abstract}

\section{Introduction}

Training-free open-vocabulary semantic segmentation (OVSS) aims to segment
arbitrary categories without optimizing a task-specific segmentation model.
Frozen vision-language models provide strong semantic alignment, but their
image-level pretraining objective does not guarantee spatially coherent dense
features. In practice, prediction confidence is often concentrated in a few
high-support regions, while ambiguous boundaries, thin parts, and small
objects remain weakly supported.

This paper studies this issue from the perspective of \emph{support density}.
A visual token is reliable for dense prediction when it is supported by a
compact set of semantically related tokens and by local spatial evidence. When
this support is sparse, direct text-image similarity can produce fragmented
masks. Therefore, rather than introducing learnable decoders, DeFT performs
training-free feature tracing on the frozen visual tokens.

DeFT contains three components. First, it estimates support density using the
top-$k$ visual-token neighborhood. Second, it performs density-guided evidence
tracing from graph neighbors and local spatial windows. Third, it applies
structure-conserving multi-scale recalibration to recover continuity in
low-support regions. The refined tokens are then compared with text embeddings
using the original cosine-similarity classifier.

\section{Related Work}

\paragraph{Open-vocabulary semantic segmentation.}
OVSS extends semantic segmentation to arbitrary language-defined categories.
Training-based methods adapt vision-language features with extra supervision,
while training-free methods reuse frozen image-text representations. DeFT
belongs to the training-free family.

\paragraph{Dense transfer from vision-language models.}
CLIP-like models learn strong image-text alignment, but their patch features
are not optimized for dense prediction. Existing methods improve dense
transfer with feature correction, prompt design, attention refinement, or
post-processing. DeFT is complementary: it operates on frozen visual tokens and
uses support density to decide where structure should be recovered.

\paragraph{Structure recovery without training.}
Graph propagation and spatial smoothing are common tools for dense prediction.
Unlike trainable graph modules, DeFT uses a fixed similarity graph and
deterministic local aggregation, preserving the training-free protocol.

\section{Methodology}

Let $X \in \mathbb{R}^{N \times C}$ be frozen visual patch tokens, where
$N=H \times W$, and let $T \in \mathbb{R}^{K \times C}$ be text embeddings for
$K$ candidate classes. DeFT refines $X$ before dense classification:
\begin{equation}
    \hat{X} = \mathrm{DeFT}(X), \qquad
    Y = \hat{X}T^\top .
\end{equation}

\subsection{Overview}

The whole framework follows the frozen-runtime inference pipeline. Given an
image, the visual encoder extracts patch tokens. DeFT refines these tokens
with support-density tracing. The refined tokens are normalized and compared
with text embeddings to obtain segmentation logits.

\subsection{Support Density Estimation}

For each normalized token $x_i$, DeFT computes cosine similarity with all
other tokens and selects its top-$k$ neighbors. The support density is the
average positive similarity over these neighbors:
\begin{equation}
    d_i = \frac{1}{k}\sum_{j \in \mathcal{N}_k(i)} \max(s_{ij}, 0).
\end{equation}
A density gate is generated by a sigmoid function so that low-density tokens
receive stronger refinement.

\subsection{Density-Guided Evidence Tracing}

DeFT gathers graph evidence from semantically related top-$k$ tokens and local
evidence from spatial neighborhoods on the feature map. The two sources are
mixed with fixed weights. A density gate determines how much each original
token should be replaced by traced evidence. High-support tokens remain close
to their original representation, while low-support tokens borrow reliable
evidence from their neighbors.

\subsection{Structure-Conserving Recalibration}

After tracing, DeFT applies deterministic multi-scale residual recalibration.
Average-pooled residuals from multiple kernel sizes are injected more strongly
into low-density regions. This step improves continuity while preserving the
original cosine-similarity interface with text embeddings.

\section{Experiment}

\subsection{Experimental Settings}

\paragraph{Datasets.}
Following the reference protocol, experiments cover PASCAL VOC, PASCAL
Context, COCO, ADE20K, and Cityscapes. The table columns are grouped into
with-background datasets (V21, PC60, Object) and without-background datasets
(V20, PC59, Stuff, ADE, City).

\paragraph{Implementation Details.}
We use a frozen CLIP ViT-B/16 runtime unless otherwise specified. Images are
resized with a short side of 336 pixels, and sliding-window inference uses a
224 $\times$ 224 crop with a 112 $\times$ 112 stride. DeFT uses fixed
hyperparameters across datasets: $k=12$, density threshold $0.46$, trace
strength $0.72$, graph weight $0.65$, local weight $0.35$, and multi-scale
kernels $\{3,5,7\}$.

\subsection{Comparison with State-of-the-Art Methods}

The main result table follows the reference paper. Fill only the final DeFT
row after running \texttt{bash scripts/run\_deft\_all8.sh}.

\begin{table*}[t]
\centering
\small
\caption{Quantitative comparison across eight segmentation benchmark datasets.
The final DeFT row is intentionally left blank for experimental results.}
\begin{tabular}{lcccccccccccc}
\toprule
Methods & Pub. & Extra & TF & V21 & PC60 & Obj. & V20 & PC59 & Stuff & ADE & City & Avg. \\
\midrule
GroupViT & CVPR'22 & $\times$ & $\times$ & 50.4 & 18.7 & 27.5 & 79.7 & 23.4 & 15.3 & 9.2 & 11.1 & 29.4 \\
SegCLIP & ICML'23 & $\times$ & $\times$ & 52.6 & 24.7 & 26.5 & - & - & - & - & - & - \\
TCL & CVPR'23 & $\times$ & $\times$ & 55.0 & 30.4 & 31.6 & 83.2 & 33.9 & 22.4 & 17.1 & 24.0 & 37.2 \\
DINOiser & ECCV'24 & DINO & $\times$ & 62.1 & 32.4 & 34.8 & 80.9 & 35.9 & 24.6 & 20.0 & 31.7 & 40.3 \\
SAM-CLIP & ECCV'24 & SAM & $\times$ & 60.6 & 29.2 & - & - & - & - & 17.1 & - & - \\
PnP-OVSS & CVPR'24 & BLIP & \checkmark & - & - & 36.2 & 51.3 & 28.0 & 17.9 & 14.2 & - & - \\
LaVG & ECCV'24 & DINO & \checkmark & 62.1 & 31.6 & 34.2 & 82.5 & 34.7 & 23.2 & 15.8 & 26.2 & 38.8 \\
ProxyCLIP & ECCV'24 & DINO & \checkmark & 61.3 & 35.3 & 37.5 & 80.3 & 39.1 & 26.5 & 20.2 & 38.1 & 42.3 \\
CLIP & ICML'21 & $\times$ & \checkmark & 18.6 & 9.9 & 8.1 & 49.4 & 11.1 & 5.7 & 3.1 & 6.5 & 14.1 \\
MaskCLIP & ECCV'22 & $\times$ & \checkmark & 38.8 & 23.6 & 20.6 & 74.9 & 26.4 & 16.4 & 9.8 & 12.6 & 27.9 \\
CLIPSurgery & PR'25 & $\times$ & \checkmark & - & 29.3 & - & - & - & 21.9 & - & 31.4 & - \\
CaR & CVPR'24 & $\times$ & \checkmark & 48.6 & 30.5 & 36.6 & 73.7 & 39.5 & - & 17.7 & - & - \\
GEM & CVPR'24 & $\times$ & \checkmark & 46.2 & - & - & - & 32.6 & 15.7 & - & - & - \\
CLIPtrase & ECCV'24 & $\times$ & \checkmark & 50.9 & 29.9 & 43.6 & 81.0 & 33.8 & 22.8 & 16.4 & 21.3 & 37.5 \\
ClearCLIP & ECCV'24 & $\times$ & \checkmark & 57.0 & 32.6 & 33.0 & 80.9 & 35.9 & 23.9 & 16.7 & 30.0 & 38.8 \\
SCLIP & ECCV'24 & $\times$ & \checkmark & 59.7 & 31.7 & 33.5 & 81.5 & 34.5 & 22.7 & 16.5 & 32.3 & 39.1 \\
NACLIP & WACV'25 & $\times$ & \checkmark & 58.9 & 32.2 & 33.2 & 79.7 & 35.2 & 23.3 & 17.4 & 35.5 & 39.4 \\
Feature-purified runtime & ICCV'25 & $\times$ & \checkmark & 63.9 & 37.2 & 37.9 & 84.5 & 39.9 & 26.4 & 20.8 & 41.1 & 44.0 \\
\textbf{DeFT (Ours)} & AAAI'27 & $\times$ & \checkmark & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} & \textbf{TBD} \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Ablation Study}

The ablation table follows the same V21/ADE/Avg. layout. Fill it after running
\texttt{bash scripts/run\_deft\_ablation\_v21\_ade.sh}.

\begin{table}[t]
\centering
\caption{Ablation results on each DeFT component.}
\begin{tabular}{cccccc}
\toprule
Graph & Local & Multi-scale & V21 & ADE & Avg. \\
\midrule
 &  &  & TBD & TBD & TBD \\
\checkmark &  &  & TBD & TBD & TBD \\
\checkmark & \checkmark &  & TBD & TBD & TBD \\
\checkmark & \checkmark & \checkmark & TBD & TBD & TBD \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Sensitivity to the number of tracing neighbors $k$.}
\begin{tabular}{lccccc}
\toprule
$k$ & 4 & 8 & 12 & 16 & 24 \\
\midrule
V21 & TBD & TBD & TBD & TBD & TBD \\
ADE & TBD & TBD & TBD & TBD & TBD \\
Avg. & TBD & TBD & TBD & TBD & TBD \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Comparison across different CLIP backbones.}
\begin{tabular}{lccc}
\toprule
Backbone & V21 & ADE & Avg. \\
\midrule
ViT-B/32 & TBD & TBD & TBD \\
ViT-B/16 & TBD & TBD & TBD \\
ViT-L/14 & TBD & TBD & TBD \\
\bottomrule
\end{tabular}
\end{table}

\section{Conclusion}

We present DeFT, a training-free density-guided feature tracing framework for
open-vocabulary semantic segmentation. DeFT addresses support density
imbalance by estimating token reliability, tracing graph-local evidence, and
applying structure-conserving recalibration. The framework preserves the
frozen vision-language runtime while improving dense spatial predictions.
After the provided scripts are run, the final tables can be filled directly.

\end{document}
